General purpose gene annotation using deep learning

### 2020-04-21 paper/main.tex
literature

\begin{methods}
\section{Methods}
We worked with 192 animal and 60 plant genomes. The data of each genome consists of the latest publically available genomic assembly in form of a FASTA file and the latest annotation in the GFF format. We used all animals genomes in EnsemblMetazoa 45 (\citealp{Howe19}) as well as all non-embargoed plant genomes of the JGI Phytosome 13 database (\citealp{Phytozome}). Both data groups were used seperately during training as well as during evaluation. At the start of our work we split both genome groups into {\it training genomes} and {\it test genomes}. The test genome group is made up of 19 and 6 genomes, respectively, and was chosen to be of the best possible quality based on collected metadata while also representing a broad phylogenetic spread. We then experimented which genomes to use for during training from the pool of non-test genomes while evaluating the performance of models on the remaining ones. We call the genomes that were ultimatily not used during training the {\it validation genomes}. While we did split off a group of test genomes at the beginning of our work, we ended up only reporting on the generalization performance of the combined set of validation and test genomes as the validation genomes are by far the largest set. We also did not find a noticible difference in performance between those two groups.

We pre-processed and stored the genomic information by using Geenuff (\citealp{Denton19}). Geenuff is a tool for checking and exploring genomic data, which stores all information inside a SQL database. Training and evaluation data was generated by querying this database and transforming the returned data into a numerical format suitable for machine learning. The encoding of the genomic sequence is done in line with the IUPAC nucleic acid notation and the functional annotation used as labels during training is transformed to a one hot encoding with the four classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}. See also Table \ref{tab:arrays} for a more detailed description of the generated data formats.

@@

\subsection{Model Architecture and Usage}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer normalization (\citealp{Layernorm}) between each BLSTM layer to produce the predictions in the form of a base pair wise classification. The model consists of circa 5.4 million parameters and was implemented with the deep learning library keras (citation) on top of TensorFlow (\citealp{Tensorflow}). We tested multiple different architectures before arriving at this model configuration, including convolutional neural networks (CNN) and hybrid architectures. The hyperparameters were determined by a combination of manual and automatic hyperparameter optimization (talk about TPE?, quote nni?). All relevant hyperparameters can be found in Table \ref{tab:params} or in the Helixer source code repository.

>>>>

\begin{methods}
\section{Methods}
We worked with 192 animal and 60 plant genomes. The data of each genome consists of the latest publically available genomic assembly in form of a FASTA file and the latest annotation in the GFF format. We used all animals genomes in EnsemblMetazoa 45 \citep{howe2020ensembl} as well as all non-embargoed plant genomes of the JGI Phytosome 13 database \citep{goodstein2012phytozome}. Both data groups were used seperately during training as well as during evaluation. At the start of our work we split both genome groups into {\it training genomes} and {\it test genomes}. The test genome group is made up of 19 and 6 genomes, respectively, and was chosen to be of the best possible quality based on collected metadata while also representing a broad phylogenetic spread. We then experimented which genomes to use for during training from the pool of non-test genomes while evaluating the performance of models on the remaining ones. We call the genomes that were ultimatily not used during training the {\it validation genomes}. While we did split off a group of test genomes at the beginning of our work, we ended up only reporting on the generalization performance of the combined set of validation and test genomes as the validation genomes are by far the largest set. We also did not find a noticible difference in performance between those two groups.

We pre-processed and stored the genomic information by using Geenuff \citep{denton2019}. Geenuff is a tool for checking and exploring genomic data, which stores all information inside a SQL database. Training and evaluation data was generated by querying this database and transforming the returned data into a numerical format suitable for machine learning. The encoding of the genomic sequence is done in line with the IUPAC nucleic acid notation and the functional annotation used as labels during training is transformed to a one hot encoding with the four classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}. See also Table \ref{tab:arrays} for a more detailed description of the generated data formats.

@@

\subsection{Model Architecture and Usage}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer normalization \citep{ba2016layer} between each BLSTM layer to produce the predictions in the form of a base pair wise classification. The model consists of circa 5.4 million parameters and was implemented with the deep learning library keras \citep{chollet2015keras} on top of TensorFlow \citep{abadi2016tensorflow}. We tested multiple different architectures before arriving at this model configuration, including convolutional neural networks (CNN) and hybrid architectures. The hyperparameters were optimized by a combination of manual and automatic optimization. Automatic optimization was carried out by using either the TPE algorithm \citep{bergstra2011algorithms}, random search or grid search depending on the situation. We used NNI \citep{nni2019} to facilitate the search. All relevant hyperparameters can be found in Table \ref{tab:params} or in the Helixer source code repository.
